Geometric signs reveal changes social structures and networks during the Western Eurasian Aurignacian

### 2025-06-09 paper/paper.qmd
added new paper citations

Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020], coinciding with the ongoing biological transition from Neanderthals to early modern humans. Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is often linked to the dispersal of _Homo sapiens_ across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper we investigate Aurignacian social organization dynamics implied by the distribution and variation of geometric sign types found on mobile objects in Europe across four phases of the Aurignacian period.

@@

One form of longer-term residential movement that may be related to non-utilitarian mobility are aggregation/dispersion patterns, which are defined by @conkley_agg as a type of hunter-gather settlement pattern in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate. Some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], including, in some cases, decorative objects [@Kelly_2024]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous work [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] has used material culture analysis to reconstruct Upper Paleolithic social networks, which have been argued to be a driver of cultural and evolutionary change [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

>>>>

Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020], coinciding with the ongoing biological transition from Neanderthals to early modern humans. Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@dErrico_Baker_Rigaud_2025; @baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is often linked to the dispersal of _Homo sapiens_ across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper we investigate Aurignacian social organization dynamics implied by the distribution and variation of geometric sign types found on mobile objects in Europe across four phases of the Aurignacian period.

@@

One form of longer-term residential movement that may be related to non-utilitarian mobility are aggregation/dispersion patterns, which are defined by @conkley_agg as a type of hunter-gather settlement pattern in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate. Some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], including, in some cases, decorative objects [@Kelly_2024]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@dErrico_Baker_Rigaud_2025; @Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous work [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] has used material culture analysis to reconstruct Upper Paleolithic social networks, which have been argued to be a driver of cultural and evolutionary change [@dErrico_Baker_Rigaud_2025; @Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
